\chapter{C compiler support}
\label{chap:c-compiler}

The CHERI version of \keyword{Clang}, the C language front end for LLVM, has been modified to expose capabilities at the C level.

\section{Supported targets}

When targeting CHERI, you must specify a \keyword{target triple} that contains \flag{cheri} in the CPU part.  The most common triple to use is \flag{cheri-unknown-freebsd}.  This triple contains three portions, in the form \textit{cpu-vendor-os}.  The vendor is irrelevant, only the CPU and OS matter for code generation.

There are two ways to pass the triple to Clang.  The first is to add a \flag{-target} flag, followed by your triple, to your compile flags.  The second is to create a symbolic link to clang of the form \textit{cheri-unknown-freebsd-clang}.  

\begin{notebox}{\file{build\_sdk.sh}}
If you have the cross-compiler from the SDK, then it will automatically have its default target triple (and sysroot) configured correctly.  You can omit the \flag{-target} option.  You can see the default target triple by passing \flag{-v} to \file{clang}.
\end{notebox}

\section{The \sandboxABI}

\index{\sandboxABI{}}
You can select the \sandboxABI{} by passing \flag{-mabi=purecap} to clang.  When compiling in this mode, \textit{all} pointers (including function pointers) will be represented as capabilities.

The compiler will infer the size of all stack allocations automatically and expects memory allocators to set the size appropriately for their allocations.

C code in this model will not contain any non-capability load or store instructions.  This means that it is safe to clear the ambient data capability (\creg{0}) when in this mode.  It could be used as a general-purpose capability register, but is not currently.  In this mode, or one where the ambient data capability is very limited in scope, this provides full memory safety.

\section{Pointer qualifiers}

\index{pointers}\index{qualifiers}
In the compatible ABI, the \ccode{__capability} qualifier indicates that a pointer should be represented by a memory capability.
For example:

\begin{csnippet}
void *a; // 64-bit integer interpreted as a pointer
void * __capability b; // 256-bit capability interpreted as a pointer
\end{csnippet}

Implicit casts between capability and integer pointers are not permitted.
This is an intentional design decision, as such casts require careful thought to ensure that they are valid.

Clang exposes a feature-test macro that allows you to easily check whether you're compiling for a target that supports capabilities and to \ccode{#define}-way the \ccode{__capability} qualifier for other platforms:

\begin{csnippet}
	#if !__has_feature(capabilities) && !defined(__capability)
	#   define __capability
	#endif
\end{csnippet}


\section{Pragmas for generating capabilities}

\index{pointer\_interpretation pragma}
Often is it useful to define a block of code where pointers will be interpreted as either capabilities or integers, rather than having to annotate them individually.
This is possible with the \ccode{pointer_interpretation} pragma.
As with other C99 pragmas, this can be used either with the \ccode{#pragma} or \ccode{_Pragma()} syntax.

The valid options for this pragma are listed in \autoref{tbl:pointerinterpretation}.
Most uses of this pragma are likely to be via macros that explicitly push and pop.
For example, the following definitions provide a mechanism for defining ranges:

\begin{table}
	\begin{center}
		\begin{tabu}{*{2}{c}}
			\toprule
			\headerrow
			Option & Meaning\\
			\midrule
			\ccode{push} & Save the current interpretation. \\
			\ccode{pop} & Restore a previously saved interpretation. \\
			\ccode{integer} & Pointers are represented by integers. \\
			\ccode{capability} & Pointers are represented by capabilities. \\
			\ccode{default} & Pointers are represented by whatever the default is for this target. \\
			\bottomrule
		\end{tabu}
		\caption{\label{tbl:pointerinterpretation}The valid options for the pointer interpretation pragma.}
	\end{center}
\end{table}

\begin{csnippet}
#if __has_feature(pointer_interpretation)
#	define BEGIN_CAPABILITIES \
		_Pragma("pointer_interpretation push") \
		_Pragma("pointer_interpretation capability")
#	define END_CAPABILITIES \
		_Pragma("pointer_interpretation pop")
#	define BEGIN_NO_CAPABILITIES \
		_Pragma("pointer_interpretation push") \
		_Pragma("pointer_interpretation integer")
#	 END_NO_CAPABILITIES
		_Pragma("pointer_interpretation pop")
#else
#	define BEGIN_CAPABILITIES
		_Pragma("GCC error \"Compiler does not support capabilities\"")
#	define END_CAPABILITIES
		_Pragma("GCC error \"Compiler does not support capabilities\"")
#	define BEGIN_NO_CAPABILITIES
#	define END_NO_CAPABILITIES
#endif
\end{csnippet}

When compiled with a compiler that supports this pragma, they will have their expected meanings (including correctly handling nesting).
When compiled with another compiler, the \ccode{NO_CAPABILITIES} versions will be silently ignored, but the \ccode{CAPABILITIES} versions will raise an error.

This allows code like this:

\begin{csnippet}
BEGIN_CAPABILITIES
struct test
{
	int *a,*b,*c;
} t;
END_CAPABILITIES
\end{csnippet}

This is equivalent to explicitly annotating each field in the \ccode{struct} with the \ccode{__capability} annotation.


\section{Built-in functions}

The C front end provides a number of built-in functions for manipulating capabilities, listed in \autoref{tbl:builtins}.
These correspond directly to a single instruction, shown in the table.
Most of these functions (those with \ccode{cheri} in their name) are generic and should apply to any architecture that supports memory capabilities.
The remainder (those with \ccode{cheri} in their name) are specific to the current implementation.

In the pure-capability mode, the \ccode{__builtin_return_address} builtin function returns the capability that will be used to return.

\subsection{Alignment-related builtins}

Our experiments have shown that a large proportion of the pointer manipulations that are incompatible with CHERI C are as a result of attempting to determine the alignment of a pointer.
To simplify porting, we have introduced new builtins that will correctly report or adjust alignment of a pointer, irrespective of its underlying representation (integer or capability).
These can be wrapped in macros for compatibility with existing compilers, though we hope that they will be widely adopted.

The following pair of builtins return whether the pointer given by the first argument is aligned:

\begin{csnippet}
_Bool __builtin_is_aligned(void *, size_t);
_Bool __builtin_is_p2aligned(void *, size_t);
\end{csnippet}

The first variant takes the desired alignment as the second parameter.
The second takes the base-2 log of the alignment.
It is undefined behaviour to provide a value that is not a power of two as the argument to the first form, the second makes it impossible to do so accidentally.

The remaining builtins adjust the alignment of a pointer, rounding the address up or down as requested:

\begin{csnippet}
void *__builtin_align_up(void *, size_t);
void *__builtin_p2align_up(void *, size_t);
void *__builtin_align_down(void *, size_t);
void *__builtin_p2align_down(void *, size_t);
\end{csnippet}

As before, each rounding direction comes with two options for specifying the alignment: either as a desired value or as the base-2 logarithm of the desired value.

\begin{table}
	\begin{center}
		\begin{tabu}{ll}
			\toprule
			\headerrow
			Clang builtin & CHERI instruction\\
			\midrule
			\ccode{__builtin_cheri_length_set} & \asm{CSetLen} \\
			\ccode{__builtin_cheri_length_get} & \asm{CGetLen} \\
			\ccode{__builtin_cheri_bounds_set} & \asm{CSetBounds} \\
			\ccode{__builtin_cheri_base_increment} & \asm{CIncBase} \\
			\ccode{__builtin_cheri_base_get} & \asm{CGetBase} \\
			\ccode{__builtin_cheri_perms_and} & \asm{CAndPerm} \\
			\ccode{__builtin_cheri_perms_get} & \asm{CGetPerm} \\
			\ccode{__builtin_cheri_type_get} & \asm{CGetType} \\
			\ccode{__builtin_cheri_tag_get } & \asm{CGetTag} \\
			\ccode{__builtin_cheri_sealed_get } & \asm{CGetSealed} \\
			\ccode{__builtin_cheri_tag_clear } & \asm{CClearTag} \\
			\ccode{__builtin_cheri_seal} & \asm{CSeal} \\
			\ccode{__builtin_cheri_unseal } & \asm{CUnseal} \\
			\ccode{__builtin_cheri_perms_check} & \asm{CCheckPerm} \\
			\ccode{__builtin_cheri_type_check} & \asm{CCheckType} \\
			\ccode{__builtin_cheri_offset_increment} & \asm{CIncOffset} \\
			\ccode{__builtin_cheri_offset_set} & \asm{CSetOffset} \\
			\ccode{__builtin_cheri_offset_get} & \asm{CGetOffset} \\
			\ccode{__builtin_cheri_program_counter_get} & \asm{CGetPCC} \\
			\ccode{__builtin_cheri_global_data_get} & \asm{CMove \%0, $c0} \\ %stopzone
			\ccode{__builtin_cheri_stack_get} & \asm{CMove \%0, $c11} \\ %stopzone
			\midrule
			\ccode{__builtin_mips_cheri_cause.get} & \asm{CGetCause} \\
			\ccode{__builtin_mips_cheri_cause.set} & \asm{CSetCause} \\
			\ccode{__builtin_mips_cheri_invoke_data_cap_get} & \asm{CMove \%0, $c26} \\ %stopzone
			\ccode{__builtin_mips_cheri_kernel_cap1_get} & \asm{CMove \%0, $c27} \\ %stopzone
			\ccode{__builtin_mips_cheri_kernel_cap1_get} & \asm{CMove \%0, $c28} \\ %stopzone
			\ccode{__builtin_mips_cheri_kernel_code_cap_get} & \asm{CMove \%0, $c29} \\ %stopzone
			\ccode{__builtin_mips_cheri_kernel_data_cap_get} & \asm{CMove \%0, $c30} \\ %stopzone
			\ccode{__builtin_mips_cheri_exception_program_counter_cap_get} & \asm{CMove \%0, $c31} \\ %stopzone
			\bottomrule
		\end{tabu}
		\caption{\label{tbl:builtins}C built-in functions provided for CHERI.}
	\end{center}
\end{table}



\section{Predefined macros}

CHERI Clang provides a number of feature test macros, shown in \autoref{tbl:featmacros}.
These are intended to be used to check for specific functionality and conditionally compile code.

\begin{table}
	\begin{center}
		\begin{tabu}{llp{5cm}}
			\toprule
			\headerrow
			Macro & Value & Meaning \\
			\midrule
			\ccode{_MIPS_ARCH} & \ccode{"cheri"} & The variant of the MIPS architecture in use. \\
			\ccode{_MIPS_ARCH_CHERI} & 1 & CHERI is the target (MIPS-specific). \\
			\ccode{__CHERI__} & 1 & CHERI is the target (may be used for non-MIPS implementations). \\
			\ccode{_MIPS_SZCAP} & 256 & The size of a capability (in bits). \\
			\ccode{__CHERI_PURE_CAPABILITY__} & 1 & Set only if targeting the \sandboxABI{}.\\
			\ccode{__CHERI_SANDBOX__} & 3 & Set only if targeting the \sandboxABI{}.\\
			\bottomrule
		\end{tabu}
		\caption{\label{tbl:featmacros} Feature test predefined macros supported by CHERI Clang}
	\end{center}
\end{table}

\sloppy
Additionally, the permission flags are provided as symbolic constants in the form of predefined macros.
These all start with \ccode{__CHERI_CAP_PERMISSION_} followed by a suffix from \autoref{tbl:permmacros}.
For example, \ccode{__CHERI_CAP_PERMISSION_PERMIT_SEAL__} is the permission bit for sealing.

\begin{table}
	\begin{center}
		\begin{tabu}{llp{4cm}}
			\toprule
			\headerrow
			Macro suffix & Value & Meaning \\
			\midrule
			\ccode{_GLOBAL__} & 1 & Global ``permission'' flag value. \\
			\ccode{_PERMIT_EXECUTE__} & 2 & Execute permission flag value. \\
			\ccode{_PERMIT_LOAD__} & 4 & Load (data) permission flag value. \\
			\ccode{_PERMIT_STORE__} & 8 & Store (data) permission flag value. \\
			\ccode{_PERMIT_LOAD_CAPABILITY__} & 16 & Load-capability permission flag value. \\
			\ccode{_PERMIT_STORE_CAPABILITY__} & 32 & Store-capability permission flag value. \\
			\ccode{_PERMIT_STORE_LOCAL__} & 64 & Store ephemeral permission flag value. \\
			\ccode{_PERMIT_SEAL__} & 128 & Seal permission flag value. \\
			\ccode{_ACCESS_EPCC__} & 1024 & Access \reg{epcc} permission flag value. \\
			\ccode{_ACCESS_KCC__} & 4096 & Access \reg{kcc} permission flag value. \\
			\ccode{_ACCESS_KDC__} & 2048 & Access \reg{kdc} permission flag value. \\
			\ccode{_ACCESS_KR1C__} & 8192 & Access \reg{kr1c} permission flag value.\\
			\ccode{_ACCESS_KR2C__} & 16384 & Access \reg{kr2c} permission flag value.\\
			\bottomrule
		\end{tabu}
		\caption{\label{tbl:permmacros} Suffixes of permission value predefined macros supported by CHERI Clang.  \ccode{__CHERI_CAP_PERMISSION_} is prefixed to all of these.}
	\end{center}
\end{table}

\section{\texttt{\_\_intcap\_t}}

CHERI provides two builtin types, \ccode{__intcap_t} and \ccode{__uintcap_t}.  These are expected to be \ccode{typedef}'d to \ccode{[u]intcap_t} in the compatible ABI and \ccode{[u]intptr_t} in the \sandboxABI.  These types are the same sizes as capabilities (i.e. \ccode{sizeof(__intcap_t) == sizeof(void* __capability)}).  

As with \ccode{intptr_t} in conventional C code, the goal for these types is to support any integer value or any pointer value, and to allow arbitrary arithmetic.  The C standard requires that, if \ccode{intptr_t} exists, it should be possible to store a pointer in it and recover the same pointer.  Most C code expects to be able to also perform arithmetic on the pointer value.

\index{null capability}
When an integer value is cast to an \ccode{__intcap_t}, the compiler will set that integer as the offset of a canonical null capability.  Any arithmetic on an \ccode{__intcap_t} value is performed by extracting its offset, manipulating it, and then setting it.  This ensures that integer arithmetic on \ccode{__intcap_t} values derived from integers will work precisely as expected.  Arithmetic on values derived from pointers (capabilities) is a little bit more complex.  Masking to access the unused low and high bits should still work as expected, as should addition.  Multiplication and division, however, will not.

Comparisons between \ccode{__intcap_t} values use special pointer comparison instructions.
These have the semantics that any untagged capability (including those derived from the null capability) will compare before those with a valid tag.
Within these two regions, capabilities are compared based on their absolute virtual addresses (i.e. base + offset).
Comparisons between signed and unsigned \ccode{__intcap_t} values will use the signed and unsigned variants of the compare instructions.

\begin{notebox}{Pointer comparison and garbage collection}
The C specification makes comparisons between pointers to different objects undefined.
The CHERI C compiler aims to ensure that such comparison is stable and has the same result as pointer comparison with non-capability pointers.

The fact that the CHERI model makes it possible to differentiate pointers from integers (even in the case of \ccode{__intcap_t}) means that it is possible to implement a copying garbage collector for CHERI C.
The compiler aims to ensure that virtual addresses do not leak into integer registers unless the programmer explicitly requests them to (for example, casting a capability to an \ccode{int} or explicitly accessing the base).
This means that, as long as care is taken, memory referenced by capabilities but outside of the global data capability range can be safely collected, unless the code relies on undefined behavior or explicitly attempts to subvert the garbage collector.
\end{notebox}

\section{Pointer casts}

In the \hybridABI{}, pointers may be represented either as integers or capabilities.
Implicit casts between the two kinds are potentially dangerous for several reasons:

\begin{itemize}
	\item A capability may extend outside of the range covered by the default data capability and so not become \ccode{NULL} as a result of the cast.
	\item A capability cast to an integer pointer will lose bounds and permissions information and so may result in memory safety errors.
	\item A capability constructed from an integer will have lack any fine-gained permissions or bounds, yet code receiving it may expect it to do so.
\end{itemize}

It is therefore important that all such casts be audited for correctness.
The compiler provides \ccode{__cheri_tocap} and \ccode{__cheri_fromcap} casts and will provide a warning on all casts that are not correctly annotated.

\section{Casts between pointers and integers}

In the hybrid ABI, most pointers are represented as integers and the cast between the two representations is purely a construct of the language type system.
In this ABI, casts between integers and \ccode{__capability}-qualified pointers are considered as being equivalent to cast between an integer-represented pointer and a capability-represented pointer.
In particular, the value of a cast should be the same independent of the intermediate integer step:
\begin{csnippet}
	void *src;
	void * __capability dest;
	void * __capability dest2;
	long mid = (long)src;
	// These are equivalent:
	dest = (void* __capability)src;
	dest2 = (void* __capability)mid;
\end{csnippet}

In the \sandboxABI, there is no guarantee within the abstract machine that it is possible to preserve a pointer via a cast to any integer type other than \ccode{__intcap_t} or \ccode{__uintcap_t}.
A cast to a primitive integer one of these types will return the offset.
The compiler supports two kinds of explicit cast that explicitly produce either the offset or the (virtual) address:

\begin{csnippet}
	void *somePointer;
	// Get the address as an integer
	long addr = (__cheri_addr long)somePointer;
	// Get the offset as an integer
	long offset = (__cheri_offset long)somePointer;
\end{csnippet}

Casts without one of these annotations will produce a warning and implicitly give the address.
Annotating every cast to a specific type is sometimes cumbersome and so the compiler also supports a \ccode{memory_address} attribute that marks an integer type as being expected to contain virtual addresses.

A pointer constructed from an integer in the \sandboxABI{} is a canonical null capability with its offset set to the source integer value.

\section{Input and output}

The CHERI capability model supports pointers with restricted rights.
An early version of CHERI Clang used this to enforce \ccode{const} in hardware.
This proved problematic, for example in idioms such as \ccode{strchr}:

\begin{csnippet}
char *strchr(const char *s, int c);
\end{csnippet}

This function returns a non-\ccode{const} pointer derived from a \ccode{const} pointer.
The invariant that this intends to document is that the \ccode{strchr} function will not modify the buffer \ccode{s}.
In the capability model, it is impossible to derive a writable capability from a read-only capability, which broke any caller of this function that expected the returned pointer to be writable.

To address this, CHERI Clang provides an \ccode{__input} qualifier, which is similar to \ccode{const} but enforced in hardware.
Capabilities can be implicitly cast to \ccode{__input}-qualified versions, including as function parameters, and the compiler will insert a \asm{CAndPerms} instruction to disclaim write permissions.

A corresponding \ccode{__output} qualifier has similar behavior, creating a write-only capability.
For example, consider the following simple snippet:

\begin{csnippet}
int in(int * __input __capability x);
int out(int * __output __capability x);

void inout(int * __capability x)
{
	in(x);
	out(x);
}
\end{csnippet}

The \ccode{inout} function will compile to the following LLVM IR:

\begin{irsnippet}
; Function Attrs: nounwind
define void @inout(i32 addrspace(200)* %x) #0 {
entry:
	%0 = bitcast i32 addrspace(200)* %x to i8 addrspace(200)*
	%1 = tail call i8 addrspace(200)* @llvm.mips.cap.perms.and(i8 addrspace(200)* %0, i64 65495)
	%2 = bitcast i8 addrspace(200)* %1 to i32 addrspace(200)*
	%call = tail call i32 @in(i32 addrspace(200)* %2) #3
	%3 = tail call i8 addrspace(200)* @llvm.mips.cap.perms.and(i8 addrspace(200)* %0, i64 65515)
	%4 = bitcast i8 addrspace(200)* %3 to i32 addrspace(200)*
	%call1 = tail call i32 @out(i32 addrspace(200)* %4) #3
	ret void
}
\end{irsnippet}

The two intrinsic calls construct two capabilities derived from the original argument.
The first removes the store and store-capability permissions.
The second removes the load and load-capability permissions.
This means that the \ccode{in} function can not use its argument to write memory and the \ccode{out} function can not use its argument to read memory.

\section{Inline assembly}

Clang supports GNU-style inline assembly.
When targeting CHERI, \ccode{"C"} can be used as a register constraint indicating that a particular operand is a capability.
For example, to access the value in \creg{0}, you might write:

\begin{csnippet}
void * __capability c0;
__asm__ __volatile__ ("cmove %0, $c0"
	: "+C"(c0) /* c0 is a capability output operand */
	: /* No input operands */
	: /* No clobbers */
	);
\end{csnippet}

Note that the \ccode{+} prefix is required to indicate that this is an output register, for input operands \ccode{"C"} is correct.

\section{\file{cheri.h}}

CHERI Clang includes a header, \file{cheri.h}, which can be included by code targeting any architecture and exposes architecture-neutral definitions of types and functions for memory capabilities.

The new types and qualifiers provided by the compiler all use identifiers that start with two underscores, because this part of the namespace is reserved for the language implementation and so should not conflict with any user code.
The \file{cheri.h} header provides more human-friendly versions of these, including \ccode{intcap_t} and \ccode{uintcap_t} types, and the \ccode{capability}, \ccode{input} and \ccode{output} qualifiers.

On architectures that don't support capabilities, the two types will be equivalent to \ccode{intptr_t} and \ccode{uintptr_t} and the qualifiers will be ignored.

The header also provides versions of the \ccode{__builtin_cheri} functions, without the \ccode{__builtin_} prefix.
The versions of these that are exposed when compiling without capability support are no-ops for the set variants and define maximum permissions and bounds for the get variants.

Only the functions that are expected to be generic across all implementations of the CHERI model (as opposed to the MIPS-based concrete implementation) are exposed by this header.

\section{Compiler assistance for cross-domain calls}
\label{sec:cccall}

\index{ccall}\index{cross-domain call}
It is important to avoid leaking rights (and other information) when calling between security contexts.  CHERI implements cross-domain calls via a special instruction (\asm{ccall}) that traps to a privileged handler (currently in the kernel) to perform the domain transition.  The handler is responsible for saving and zeroing all callee-save registers and clearing all caller-save registers.  On return, the same code in the handler will restore all callee-save registers and (again) zero caller-save registers.

The handler does not know precisely what the arguments are for any function that is involved in a cross-domain call and so can not clear the unused argument registers on call, nor the unused return registers on return.

On the other side of the cross-domain call, the code must determine which method to invoke.  This is done by means of a method number, which is passed in a register that is normally unused for calls.

If a function is annotated with \ccode{__attribute__((cheri_ccall))} then the front end will replace calls to it with specially crafted calls to \ccode{cheriinvoke}, with the method number (identified by a global variable that is initialised by the sandbox loader) in the correct register.
This annotation provides two declarations of the function, one with a suffix (specified by the \ccode{cheri_method_suffix} attribute) that takes an explicit CHERI class argument and one that automatically sets the class (code and data capability) arguments from a global variable specified by the \ccode{cheri_method_class} attribute.

Functions with the \ccode{cheri_ccallee} attribute will use the ccall calling convention, and so the compiler will zero any return registers.  This is intended to allow functions to be declared that are usable both inside and between compartments.

Functions with the ccall calling convention have two extra capability argument registers, \creg{1} and \creg{2}, which contain the code and data capabilities.  
They also have an extra integer argument register (\reg{v0}), which contains the method number.  These are not exposed to functions marked as using the ccallee calling convention, which accept the normal calling convention's argument registers.  

Consider the following simple program:

\begin{csnippet}
void * __capability data;
__attribute__((cheri_ccallee))
void * __capability cgetdata(void)
{
	return data;
}

__attribute__((cheri_ccallee))
int cgetnumber(void)
{
	return 42;
}
\end{csnippet}

This contains two simple functions that are expected to be invoked as part of a domain transition.  The first returns a capability from inside its data capability, the second returns a constant.  In the first, the return value is a capability and so will be returned in capability register \creg{3}.  The two integer return registers (\reg{2} and \reg{3}, sometimes called \reg{v0} and \reg{v1}) are unused.  The compiler will generate this code for the return:

\begin{asmsnippet}
	clc     $c3, $zero, 0($c2)
	move    $2, $zero
	move    $3, $zero
\end{asmsnippet}%stopzone

The return value is loaded into \creg{3}.  The two integer return values are zeroed.  In the second function, one of the integer return registers is used, but the capability registers are not.  The compiler generates this code:

\begin{asmsnippet}
	cfromptr  $c3, $c0, $zero
	addiu     $2, $zero, 42
	move      $3, $zero
\end{asmsnippet}%stopzone

In this case, the first integer return register is set to 42, but the other return registers are cleared.  

\subsection{Cross-domain function pointers}

Cross-domain calls require more than a single capability: they require the code and data capabilities for the target, as well as the method number.
Function pointers annotated with \ccode{__attribute__((cheri_ccallback))} in the compiler will be translated to a hidden structure that contains these three components.
A call to this function pointer will be replaced by a call to \ccode{cheriinvoke} with the correct arguments set up for the cross-domain call.

Because such function pointers cannot be constructed from a simple cast, CHERI clang provides a builtin to construct them:
\begin{csnippet}
__builtin_cheri_callback_create(const char *class_name,
                                struct cheri_object object,
                                function fn);
\end{csnippet}

The final argument type is any function declared with the annotations required for exported functions.
The class name is used in combination with the function to look up the method number and the CHERI object is used directly.
This combination allows sandboxed code that is passed a function pointer of this form to call back into the specified object.

\section{C++}

CHERI C++ extends the CHERI model to the C++ language.

\subsection{Reference qualifiers}

In the compatible ABI, the \ccode{__capability} qualifier indicates that a C++ reference should be represented by a memory capability.
It also indicates that it can be initialised with the dereferenced value of a capability pointer without the need for an explicit cast. For example:

\begin{cxxsnippet}
int * __capability x;
int & __capability y = *x;
\end{cxxsnippet}

\section{Additional warnings}

The CHERI C compiler provides additional warnings to help catch assumptions that may not be valid in CHERI C.

\subsection{Capability to integer casts}

In most C implementations, it is safe to cast a pointer to a \ccode{long} or \ccode{long long} and back again.
In CHERI, this is not the case, though it is safe to cast a pointer to a \ccode{__uintcap_t} or \ccode{__intcap_t}.

Casts from pointer to integers are common and only some of them are unsafe.
The \flag{-Wcapability-to-integer-cast} flag enables warnings on all such casts, unless the integer type is qualified with the \ccode{memory_address} attribute or the cast is an explicit CHERI cast.
This warning generates a lot of false positives and so is off by default but provides a good place to start looking for problems.

\subsection{Capability misuse}

The \flag{-Wcheri-capability-misuse} flag enables a miscellaneous set of warnings related to common problems with capabilities in C.
Casts from non-capability integer types to capability types will trigger a warning, because the resulting pointer may not be dereferenced.

This flag also enables several warnings related to alignment.
For example, packed structures containing capabilities and arrays containing capabilities will trigger a warning if the resulting alignment would be unsafe to store a capability.

When performing a binary operation involving capabilities, the result will be a capability derived from the left operand.
Any binary bitwise operations performed between a capability type and a non-capability type will warn, because the result of such operations is usually expected to be a non-capability type.
The compiler can produce even more warnings on bitwise operations with the \flag{-Wcheri-bitwise-operations} flag.

\subsection{No function prototype}

If a function is defined without a prototype then, for K\&R C compatibility, it is assumed to be called with the variadic calling convention.
On most platforms, this is safe for functions with small numbers of parameters.
On CHERI-MIPS, this is not the case.

The \mipsABI{} defines a variadic calling convention in which the first 64 bytes of arguments are passed in integer registers irrespective of type.
They are then spilled to the stack in the callee.
Such a calling convention is incompatible with capabilities.
Whereas passing a floating point value in an integer register is expensive on modern pipelines, passing a capability in integer registers is impossible.
As such, if any function is declared with no prototype and called in the \sandboxABI{}, the compiler will generate invalid code.

The \flag{-Wmips-cheri-prototypes} flag instructs the compiler to warn about possible problems.

\subsection{Excess padding}

When porting code to CHERI, the size of the pointer (at least) doubles.
This can cause structures to gain additional alignment padding.
For example, a structure containing two \ccode{int}s, a \ccode{void*} and an \ccode{int} will be 16 bytes on a ILP32 platform, 24 bytes on an LP64 platform, and 48 bytes on CHERI.
Rearranging the structure to put all of the \ccode{int}s together will reduce it to 32 bytes.
If a program allocates a large number of such structures, then this can have a large impact on performance.

The \flag{-Wexcess-padding} flag makes it easier to diagnose problems related to padding.
It will report on any structure that contains a large amount of padding.
Note that this warning is not CHERI specific and can also be useful in finding structures that originated on 32-bit platforms and gained padding when moved to 64-bit systems.

This warning if off by default because it will report a lot of structures with unavoidable padding and structures whose padding has no impact on overall performance.
If code ported to CHERI is seeing a large increase in memory usage then we recommend enabling this warning to see if there are any simple fixes.

\subsection{Compiler bug warnings}

The CHERI MIPS compiler is a research prototype and may contain known bugs.
The \flag{-Wmips-cheri-bugs} flag attempts to warn on code that will trigger bugs.
This currently warns about a code generation bug when using the atomic builtins at -O0.

